Colonization-extinction dynamics determine tree species range limits in northeastern North America

### 2015-06-24 tex/manuscript.tex
Reorganized sdm scripts
Added some text to the ms and polished up script 4

%\section*{Introduction}

>>>>

\section*{Introduction}
\subsection*{Objectives}
\begin{itemize}
	\item Describe the range limits at equilibrium of ecologically important trees in eastern North America as predicted by metapopulation theory
	\item Evaluate the relationship between macroclimatic variables and colonization and extinction rates
	\item Test quantitatively whether the equilibrium prediction matches the contemporary observed range
	\item Compare among species the predicted resilience to environmental change 
\end{itemize}

Present and future rapidly changing global environmental conditions are likely to push many species to the limits of their ecological niches.
Biogeographical theory and empirical work suggests that species can respond to these changes by adapting \emph{in situ} to the new conditions, geographically tracking ideal conditions (resulting in range shifts), and/or experiencing local extinction where conditions no longer support presence.
Species distribution models (SDMs) are commonly-applied to understand the latter two options.
Classical SDMs often use a bioclimatic modeling approach that statistically relates presence or absence of the species to climate.
Such models are quite powerful when describing the present realized niche of a species, however they suffer from several drawbacks when predicting the future distribution of species, and they provide no insight into the mechanisms underlying a species' response to the environment.
An alternative more mechanistic approach is to examine the colonization rate $c$ of unoccupied patches and the extinction rate $e$ of occupied patches.
(here maybe briefly review the literature on the subject)
In such a stochastic system, the long-term expectation is that presence will only be possible where $c > e$.
It follows that, if these rates vary along an environmental gradient, the equilibrium position of the range limit on the gradient will be the point at which $c = e$.

We designed a colonization-extinction model with a dynamic parameterization that allowed the colonization and extinction rates to vary as a function of climate.
A major challenge to employing these models is the lack of time series data needed to estimate the parameters.
We used a large database of permanent forest plots, containing repeated observations of xxx plots in eastern North America (n=xxx paired observations). 
We selected the 21 most abundant species (based on the number of observations in the database) as the focus of this study.
We performed model selection, using 6 macroclimatic variables, then parameterized the model using Markov Chain Monte Carlo (MCMC) sampling and addressed the following questions for each selected species:
\begin{enumerate}
	\item Are range limits driven by colonization-extinction dynamics?
	\item To what extent are the species studied presently out of equilibrium with the environment?
	\item How resilient are range limits to perturbations in the environment?
	\item Which climatic variables best describe colonization and extinction rates?
\end{enumerate}

%% here expand a bit on each question above and perhaps provide a testable hypothesis.
We evaluated the model by reserving half of the observed sample pairs.
We parameterized the model and performed model selection on the calibration half, then used the parameters to predict the second state as a function of the climate and the first state in the evaluation dataset.
We then compared the predicted to the observed second state and computed TSS and ROC (explanation and citation?).

% can this apply to other organisms? animals prob not - dispersal is just too high
% this deserves a paragraph here
